---
title: "league_seasons"
output: html_document
---

538's [Soccer Power Index (SPI)](https://projects.fivethirtyeight.com/soccer-predictions/) rates the quality of soccer teams from a variety of leagues around the world. In this post I'll use `gganimate` to animate team SPI over the past 3 seasons.

The SPI data is available on 538's [GitHub repo](https://github.com/fivethirtyeight/data/tree/master/soccer-spi).
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

Set up the environment:
```{r}
library(tidyverse)
library(purrr)
library(gganimate)
library(ggrepel)
library(broom)
library(lubridate)

theme_set(theme_minimal(base_size = 18))
```

Load the data and make the data long, instead of having different columns for home and away results:
```{r}
data <- read_csv("data/spi_matches.csv")

df_home <- data %>% 
  select(team1, date, league, spi1) %>% 
  rename(team = team1,
         spi = spi1) %>% 
  mutate(venue = "home")

df_away <- data %>% 
  select(team2, date, league, spi2) %>% 
  rename(team = team2,
         spi = spi2) %>% 
  mutate(venue = "away")

df_all <- bind_rows(df_home, df_away) %>% 
  arrange(league, team, date) %>% 
  group_by(league, team) %>% 
  mutate(team_game_number = dense_rank(date)) %>% 
  ungroup()
```

Detect when a league's season changes
```{r}
df_all <- df_all %>% 
  filter(date <= Sys.Date())

df_league_season_dates <- df_all %>% 
  distinct(league, date) %>% 
  arrange(league, date) %>% 
  group_by(league) %>% 
  mutate(date_lag = lag(date, n = 1),
         gap = date - date_lag) %>% 
  fill(date_lag, .direction = "up") %>% 
  replace_na(list(gap = 0)) %>% 
  ungroup()

league_gaps <- df_league_season_dates %>% 
  arrange(league, desc(gap)) %>% 
  group_by(league) %>% 
  mutate(gap_number = row_number()) %>% 
  top_n(2, gap) %>% 
  ungroup() %>% 
  select(league, gap_number, gap) %>% 
  pivot_wider(names_from = gap_number, names_prefix = "gap_", values_from = gap) %>% 
  select(league, gap_1, gap_2)

df_league_season_dates <- df_league_season_dates %>% 
  left_join(league_gaps) %>% 
  #mutate(functional_gap = case_when(str_detect(league, "Austria|Russia") ~ next_gap,
  #                                  !str_detect(league, "Austria|Russia") ~ max_gap)) %>% 
  mutate(season_flag = case_when(gap >= gap_1 | gap >= gap_2  ~ TRUE,
                                 gap < gap_2 ~ FALSE)) %>% 
  replace_na(list(season_flag = FALSE))

df_league_season_dates <- df_league_season_dates %>% 
  group_by(league) %>% 
  mutate(season = cumsum(season_flag),
         season = season + 1) %>% 
  #mutate(season = case_when(functional_gap < 30 ~ 1,
  #                          functional_gap >= 30 ~ season)) %>% 
  select(league, date, season)
```
```{r}
df_league_calendar <- df_all %>% 
  left_join(df_league_season_dates) %>% 
  group_by(league) %>% 
  mutate(league_spi = median(spi)) %>% 
  ungroup() %>% 
  mutate(league = fct_reorder(league, league_spi)) %>% 
  count(league, season, date) %>% 
  complete(league, date = seq(from = min(df_all$date), to = max(df_all$date), by = "day")) %>% 
  replace_na(list(n = 0)) %>% 
  group_by(league) %>% 
  mutate(pct_games = n / sum(n))
```
```{r}
df_league_calendar %>% 
  group_by(league) %>% 
  summarize(number_of_seasons = max(season, na.rm = TRUE)) %>% 
  arrange(desc(number_of_seasons))
```
```{r}
df_league_calendar %>% 
  ggplot(aes(date, league, fill = as.factor(season))) +
    geom_tile() +
    scale_fill_viridis_d() +
    scale_x_date(expand = c(0,0)) +
    scale_y_discrete(expand = c(0,0)) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
```